Exploatarea Software și Mai Departe: Dezvăluirea Vectorilor de Atac și Contra-Măsurilor în Calculul Modern / Software Exploitation and Beyond: Unveiling Attack Vectors and Countermeasures in Modern Computing

### 2023-06-26 paper.tex
mememeememme

\documentclass{article}
\usepackage{tikz}
\usetikzlibrary{matrix,shapes,arrows,positioning,decorations.pathreplacing,arrows.meta}
\usepackage{comment}
\usepackage{marginnote}
\usepackage{pgf}
\usepackage{xstring}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

\usepackage[backend=biber,style=numeric]{biblatex}
\addbibresource{references.bib} % Import the bibliography file

\lstset{
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue}\ttfamily,
    stringstyle=\color{red}\ttfamily,
    commentstyle=\color{gray}\ttfamily,
    morecomment=[l][\color{magenta}]{\#},
    breaklines=true
}

@@

\title{Exploatarea Software și Mai Departe: Dezvăluirea Vectorilor de Atac și Contra-Măsurilor în Calculul Modern / Software Exploitation and Beyond: Unveiling Attack Vectors and Countermeasures in Modern Computing}
\maketitle
\section{Abstract}%

@@

\subsubsection{Improper input sanitization}%
In most scenarios, the fundamental mistake the software developers make is the improper validation of user-controlled input. For example, let us take the venerable Buffer Overflow vulnerability, in which an attacker can send more data than the application has space to process, resulting in Undefined Behaviour. Depending on the scenario, this undefined behavior can be exploited to redirect the execution flow of the program into code sent by the attacker (in the case of a code injection attack on the stack) or could lead to corrupting some necessary metadata somewhere else in the program (overwriting a password variable to one you control, allowing you to login without knowing the actual password). These simple techniques have mainly been mitigated in modern systems by employing passive mitigation techniques such as stack canaries and shadow stacks (which will be discussed later).

\subsubsection{Dependencies}%
Someone could be the best programmer in the world, always writing code that stands up to the highest security standards, but if they use a dependency that is vulnerable to some exploit, their program could still be at high risk. In modern software, having a programming language without a package manager is a show-stopper. Package managers have allowed developers to share and reuse code more efficiently, increasing productivity and democratization of writing excellent and valuable software. However, on the other side, the ease of using someone else's code led us to balloon our dependency graph like never before. Running someone else's code poses a security risk not only from the exploitation standpoint but also from a supply chain attack standpoint. Only one of the dependencies a project uses must be compromised for the whole application to be compromised.

@@

\subsubsection{Mixing control data with user data}
The stack is a fundamental data structure in modern computing systems, and its use in x64 systems is exciting due to the specific features and operations enabled by the x64 architecture. The stack is typically organized as a contiguous memory region with two key pointers: the stack base pointer (RBP in x64) and the stack pointer (RSP in x64). The stack grows `downward' in memory, meaning that as new items are pushed onto the stack, the stack pointer decreases in value. A typical function call on an x64 system involves pushing the return address onto the stack, followed by the function's parameters \footnote{In the most common calling convention, the first few parameters are passed in registers, not through the stack}, and finally, the local variables. These elements are popped from the stack in reverse order upon function return, ensuring correct program execution. However, the direct memory access afforded by the stack also presents a potential vector for exploitation, with techniques such as buffer overflow attacks targeting this fundamental structure. Consequently, understanding the stack's functioning in x64 systems is crucial for building and securing software systems.

\begin{figure}[ht]%
  \centering
  \begin{tikzpicture}[node distance=2cm]
    \tikzstyle{block} = [rectangle, draw, fill=blue!20, text width=5em, text
    centered, minimum height=2em, text width=3cm,
               text depth=0.25cm,
               text height=1em,
               align=center]
    \tikzstyle{line} = [draw, -latex']

@@

An example state of the stack during the call~\footnote{This example is used for simplicity. In real world scenarios the variables would have been constant propagated and the function would have directly returned the result of the computation} to the function `example`~\ref{code:example_func} is shown in~\ref{fig:stackfame-example}.

\begin{lstlisting}[
      ,label={lst:examplefn}
      ,language=C]
  int example()
  {
    int foo = 3;
    int bar = 7;

    return foo + bar;
  }
\end{lstlisting}%
\label{code:example_func}

@@

Now that we understand how a stack frame looks, we can move on to how stack frames are linked together. This\ref{fig:stacksample} example below shows the state of the stack while executing the following nested function call chain~\footnote{Assuming that foo does not return} (the `calls' operation is denoted by '$\Rightarrow$`):

@@

\subsubsection{Anatomy of the Heap}
Here we will take a small departure
from the world of semi-standardized calling conventions and venture in the
implementation defined lands of \emph{malloc} and \emph{free}. Ever since it's
inception, this api has been the bane of every C programmer's existence when
they started out programming. We will now examine in depth this API.\@

\subsubsection{malloc}
TODO add ref to standards
The malloc function is a part of the C Standard Library, as specified by both
the ANSI C and ISO C standards. It is used for dynamic memory allocation in C
programs, allowing the programmer to request a specific amount of memory space
at runtime. The malloc function, defined in the stdlib.h header file, takes a
single argument: the number of bytes to be allocated, and it returns a pointer
to the first byte of the allocated space.

The behavior of malloc is strictly defined by the C standard. Upon successful
execution, malloc provides a block of memory that is suitably aligned for any
kind of variable. The allocated memory is uninitialized, meaning it contains
indeterminate values. If the function fails to allocate the requested memory,
perhaps due to insufficient memory space, it returns a NULL pointer.
Importantly, the standard guarantees that malloc will not modify the content of
the allocated memory, which differentiates it from similar functions like
calloc, which sets the allocated memory to zero. The pointer returned by malloc
is always a unique address that is not being used elsewhere in the program,
until it is freed by the \emph{free} function.

\subsubsection{free}
Thre \emph{free} function pretty much does the inverse of what malloc does. It
marks the memory allocation starting at the address given as a parameter as free
for use by the allocator. The algorithm behind the dynamic memory allocation API
is left to be implementation defined. We will dig deeper into the
allocator used by LibC on the GNU/Linux operating system.

\subsection{Allocating memory in Linux}
The allocator used by the C standard library on linux is a variation of
\emph{ptmalloc}. There are two main ways to allocate memory using this
allocator. The allocator automatically chooses which technique to use based on
the size of the requested allocation\footnote{We will be referring to
allocations with a size $<$ {M\_MMAP\_THRESHOLD} as \emph{small} and to
allocations larger than that as \emph{big}}\footnote{{M\_MMAP\_THRESHOLD} is set
to 127k bytes by default}:
\begin{itemize}
  \item \emph{small}~-~served by the allocation algorithm
  \item \emph{large}~-~served by a call to \emph{mmap}
\end{itemize}

Large allocations bypass the allocation algorithm entirely (and will not be the
focus for our exploitation scenarios), unlike small allocations which pass
trough a pretty complex set of steps before the memory is actually given to us.
We will now discuss this algorithm in detail, pointing out areas of interest
that will help us develop exploits and mitigations later.

\subsubsection{GLibc allocation algorithm}
\begin{itemize}
  \item \emph{Arena}~-~An arena is a structure accessible by one or multiple
threads, holding references to one or several heaps. It also includes linked
lists of unallocated `chunks' within these heaps. Threads associated with a
specific arena draw their memory from the free lists of that arena.
  \item \emph{Heap}~-~A heap is a sequential memory region divided into smaller
units, known as chunks, for allocation purposes. Each heap is uniquely
associated with a single arena.
  \item \emph{Chunk}~-~A chunk represents a small, allocatable memory section.
This memory can be allocated (controlled by the application), freed (controlled
by glibc), or merged with nearby chunks to create larger sections. Importantly,
a chunk essentially serves as a shell encompassing the block of memory delivered
to the application. Every chunk is part of one heap and linked to a single
arena.
  \item \emph{Memory}~-~Memory refers to a segment of the application's address
space, generally supported by RAM or swap
\end{itemize}

@@

\item Part of the "top" chunk is split off, with a possibility of enlarging the
"top" chunk prior to this.

\item For requests requiring over-aligned \texttt{malloc}, such as
\texttt{valloc}, \texttt{pvalloc}, or \texttt{memalign}, an overly-large chunk
is located using the above \texttt{malloc} algorithm. This chunk is then divided
into two or more chunks such that the majority of the chunk is suitably aligned
and returned to the caller. The excess portions before and after this chunk are
returned to the unsorted list for future usage.
\end{enumerate}

@@

\subsubsection{Detecting heap corruption}
The \texttt{malloc} subsystem implements a series of measures aimed at detecting
heap corruption throughout its codebase. Certain checks can consistently
identify errors (for instance, passing an insufficiently aligned pointer to
\texttt{free()}). However, the majority of these checks are heuristic and may
not identify contrived, false chunks that mimic legitimate ones (for instance,
checks for forward/backward linking of chunks). Consequently, heap corruption
may persist undetected for an extended duration, or might not be reported
altogether.

Common forms of corruption are managed via calls to \texttt{malloc\_printerr};
these checks are invariably included in the code. Additional checks utilize
\texttt{assert} and can thus be disabled by building \texttt{glibc} with the
\texttt{-DNDEBUG} flag. In the current \texttt{glibc} version, both types of
checks terminate the process via a call to \texttt{\_\_libc\_message}, which
ultimately invokes \texttt{abort}. Note that although very old versions of
\texttt{glibc} offered the ability to continue execution in the presence of heap
corruption, this feature has since been discontinued.

In the following chapter we will be exploring ways to the bypass corruption checks
in order to achieve arbitrary code execution.

@@

\section{Software Exploitation Techniques}%
Software exploitation has been a prevalent issue almost as early as the
inception of software itself, underpinning the need for constant vigilance and
innovation in the field of software security. Early instances of software
exploitation can be traced back to the late 1980s with the advent of the Morris
Worm. Considered as one of the first computer worms distributed via the
Internet, the Morris Worm infected approximately 6,000 computers, causing
significant slowdowns or even rendering them unusable
\cite{spafford1989internet}. Since then, various forms of software
vulnerabilities have been exploited, including buffer overflows, injection
attacks, and privilege escalation, leading to massive data breaches, disruptions
in services, and significant financial losses \cite{owasp2017}. The history of
software exploitation elucidates the perpetual arms race between attackers
seeking to exploit vulnerabilities and defenders striving to secure systems, a
dynamic that continues to define the software security landscape today.

\subsection{Early stage}
\subsubsection{Shellcoding}
\begin{comment}
  shellcoding
  reference smashing the stack for fun and profit
\end{comment}
One of the earliest vulnerability that was exploited was the venerable bufffer
overflow. A Buffer overflow represents a crucial class of software
vulnerabilities that can serve as a launching pad for a myriad of complex
exploits, one of which includes shellcode injection. This vulnerability arises
when more data is written to a buffer than it can hold, thereby allowing an
attacker to overwrite adjacent memory locations \cite{seacord2013}. If exploited
judiciously, this can lead to the alteration of program's control flow, enabling
an attacker to execute arbitrary code. One manifestation of this is shellcoding,
where an attacker injects a sequence of bytes, often encoding command-line shell
instructions, into the memory and manipulates the program's execution flow to
execute this code \cite{one1996smashing}. This capability to execute arbitrary
code gives the attacker full control over the compromised system, thereby
illustrating the severity and implications of buffer overflow vulnerabilities.

@@

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Buffer overflow shellcoding example}
      ,label={lst:bof}
      ,language=C]
#include <stdio.h>
#include <stdlib.h>
#include <string.h>

        return 0;
}
\end{lstlisting}
\end{minipage}

@@

But running the program with the following malicious crafted input we get
a vastly different result:

\begin{lstlisting}[caption={Malicious input}
      ,label={lst:bof_malicious}
      ,language=bash]
# The `cat` command is used just to keep
# the stdin of the `say_hi` program open
$ cat shellcode_malicious_input.bin - | ./say_hi
aaaa
Hi !
ls
Makefile  say_hi  shellcode_malicious_input.bin
\end{lstlisting}

\begin{lstlisting}[caption={Hexdump of malicious input}
      ,label={lst:bof_malicous_hex}
      ,language=bash]
0000000 0000 0000 0000 0000 0000 0000 0000 0000
*
0000100 0000 0000 0000 0000 115a 0040 0000 0000
0000110 9090 9090 9090 9090 9090 9090 9090 9090
*
0000310 8148 80ec 0000 6a00 4868 2fb8 6962 2f6e
0000320 2f2f 5073 8948 68e7 6972 0101 3481 0124
0000330 0101 3101 56f6 086a 485e e601 4856 e689
0000340 d231 3b6a 0f58 0005
0000347
\end{lstlisting}
Our simple program seems to be behaving like a system shell. We will now analyse
what happened during the execution of our program. Here~\ref{lst:bof_malicous_hex}
is a hexdump of the input. Let's break it down:
\begin{enumerate}
  \item The first 0x100 bytes of the payload are just null bytes\footnote{We are
  allowed to do this because \emph{gets} does not stop reading the string when it
  encounters a \emph{NULL} (0) byte, even though strings in C are almost always
  NULL terminated.~\emph{scanf} will only return when it encounters an error or
  reads a line ending}. We do this in order to fill up the buffer that the
  application uses. These values could realistically be anything (besides line
  endings).
  \item As shown in\ref{veziacistackuala} we now need to overflow the
  \emph{base pointer}. In this exploitation scenario this can also be any value,
  but it can allow us to pivot the stack into some memory address we know in some cases.
  \item Now we overwrite the \emph{return address} to point to the address of the \emph{jump rsp} instruction.
  \item After the return address we place a \emph{nop sled}. The purpose of the nop
  sled is to allow us some leeway\footnote{When jumping into shellcode written to the stack the addresses the stack register and instruction pointer point to are close together. Stack manipulations (such as \emph{push}es and \emph{pop}s) can overwrite the shellcode, causing it to misbehave or making the processor execute illegal instructions.} when crafting the actual malicious payload.
  \item Finally we place the instructions we want to execute. The payload exemplified\ref{vezipayload} here uses the \emph{execve} syscall to replace the current running
  process with the \emph{sh} shell process.
\end{enumerate}

This traditional exploitation technique of buffer overflow into shellcode has
been effectively mitigated by a plethora of defensive mechanisms that we will
discuss in subsequent chapters, such as Address Space Layout Randomization
(ASLR), Data Execution Prevention (DEP), Stack Canaries, and Control-Flow
Integrity (CFI). Despite these advancements, the described exploitation
scenarios retain relevance in contexts where these protection measures are
absent or cannot be implemented. For example, systems lacking a Memory
Management Unit (MMU), many embedded systems, and certain router configurations
may not have these defenses, leaving them susceptible to such attacks. Thus,
understanding these exploits remains critical in a comprehensive approach to
system security.

@@

\subsubsection{Command injection}
Command injection vulnerabilities represent a pervasive class of security issues
that impact both web and system programming environments. These vulnerabilities
occur when an application passes unsafe user-supplied data (forms, cookies, HTTP
headers, etc.) to a system shell, allowing the attacker to execute arbitrary
commands on the host operating system. In the context of web applications, these
flaws are primarily found in scripts that fail to properly sanitize user input,
making them susceptible to cross-site scripting (XSS) and SQL injection attacks.
However, command injection vulnerabilities are not confined to web programming;
they are equally prevalent in systems programming. Poorly designed system
applications and scripts that accept untrusted input for command execution can
expose the system to similar attacks. This vulnerability, thus, underlines the
importance of careful data handling and proper input validation, irrespective of
the programming environment.

>>>>

\documentclass{article}
\usepackage{tikz}
\usetikzlibrary{matrix,shapes,arrows,positioning,decorations.pathreplacing,arrows.meta}
\usepackage{comment}
\usepackage{marginnote}
\usepackage{pgf}
\usepackage{xstring}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[margin=2.54cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{float}
\usepackage[toc,page]{appendix}

\usepackage[backend=biber,style=numeric]{biblatex}

\addbibresource{references.bib} % Import the bibliography file

\setcounter{secnumdepth}{4} % how many sectioning levels to assign numbers to
\setcounter{tocdepth}{4}    % how many sectioning levels to show in ToC
\lstset{
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue}\ttfamily,
    stringstyle=\color{red}\ttfamily,
    commentstyle=\color{gray}\ttfamily,
    morecomment=[l][\color{magenta}]{\#},
    breaklines=true
}

@@

\title{Exploatarea Software și Mai Departe: Dezvăluirea Vectorilor de Atac și Contra-Măsurilor în Calculul Modern / Software Exploitation and Beyond: Unveiling Attack Vectors and Countermeasures in Modern Computing}
\maketitle

\tableofcontents

\section{Abstract}%

@@

\subsubsection{Improper input sanitization}%
In most scenarios, the fundamental mistake the software developers make is the improper validation of user-controlled input. For example, let us take the venerable Buffer Overflow vulnerability, in which an attacker can send more data than the application has space to process, resulting in Undefined Behaviour.

Depending on the scenario, this undefined behavior can be exploited to redirect the execution flow of the program into code sent by the attacker (in the case of a code injection attack on the stack) or could lead to corrupting some necessary metadata somewhere else in the program (overwriting a password variable to one you control, allowing you to login without knowing the actual password). These simple techniques have mainly been mitigated in modern systems by employing passive mitigation techniques such as stack canaries and shadow stacks (which will be discussed later).

\subsubsection{Dependencies}%
Someone could be the best programmer in the world, always writing code that stands up to the highest security standards, but if they use a dependency that is vulnerable to some exploit, their program could still be at high risk. In modern software, having a programming language without a package manager is a show-stopper.

Package managers have allowed developers to share and reuse code more efficiently, increasing productivity and democratization of writing excellent and valuable software. However, on the other side, the ease of using someone else's code led us to balloon our dependency graph like never before. Running someone else's code poses a security risk not only from the exploitation standpoint but also from a supply chain attack standpoint. Only one of the dependencies a project uses must be compromised for the whole application to be compromised.

@@

\subsection{Mixing control data with user data}
The stack is a fundamental data structure in modern computing systems, and its use in x64 systems is exciting due to the specific features and operations enabled by the x64 architecture. The stack is typically organized as a contiguous memory region with two key pointers: the stack base pointer (RBP in x64) and the stack pointer (RSP in x64). The stack grows `downward' in memory, meaning that as new items are pushed onto the stack, the stack pointer decreases in value. A typical function call on an x64 system involves pushing the return address onto the stack, followed by the function's parameters \footnote{In the most common calling convention, the first few parameters are passed in registers, not through the stack}, and finally, the local variables. These elements are popped from the stack in reverse order upon function return, ensuring correct program execution. However, the direct memory access afforded by the stack also presents a potential vector for exploitation, with techniques such as buffer overflow attacks targeting this fundamental structure. Consequently, understanding the stack's functioning in x64 systems is crucial for building and securing software systems.

\begin{figure}[H]%
  \centering
  \begin{tikzpicture}[node distance=2cm]
    \tikzstyle{block} = [rectangle, draw, fill=blue!20, text width=5em, text
    centered, minimum height=2em, text width=3cm,
               text depth=0.25cm,
               text height=1em,
               align=center]
    \tikzstyle{line} = [draw, -latex']

@@

\begin{lstlisting}[
      caption={Example function},
      ,label={lst:examplefn}
      ,language=C]
  int example()
  {
    int foo = 3;
    int bar = 7;

    return foo + bar;
  }
\end{lstlisting}%

An example state of the stack during the call~\footnote{This example is used for simplicity. In real world scenarios the variables would have been constant propagated and the function would have directly returned the result of the computation} to the function `example` Listing~\ref{lst:examplefn} is shown in Figure~\ref{fig:stackfame-example}.

@@

Now that we understand how a stack frame looks, we can move on to how stack frames are linked together. Figure~\ref{fig:stacksample} shows the state of the stack while executing the following nested function call chain~\footnote{Assuming that foo does not return} (the `calls' operation is denoted by '$\Rightarrow$`):

@@

On the \emph{x86-64} platform, returning from a function can be usually abstracted to this algorithm:
\begin{enumerate}
  \item Move the current \emph{RPB} into \emph{RSP} (free-ing the current stack frame)
  \item Pop the value of \emph{Old RBP} from the stack into \emph{RBP} (restoring the caller's stack frame)
  \item Pop the \emph{Return address} into the instruction pointer (continuing execution)
\end{enumerate}

It is important to state again that in vulnerable programs, it is possible to modify the value of the return address on the stack due to improper input validation and bounds checking.

Another interesting fact about the Stack is that compilers are usually reluctant to put variables whose size is not known at compile time on the stack\footnote{Ways to achieve this exist, such as VLA's (Variable Length Arrays) or standard functions such as \emph{alloca}, but their uses is discouraged\cite{TorvaldsVLA}}. The data allocated on the Stack also only lives for as long as the function it was allocated (or any other callee of that function) is being executed.

\subsection{Anatomy of the Heap}
The Heap does not suffer from the same limitations as the Stack. Allocations made on the Heap can be programmatically free'd, instead of being managed by the compiler. The API for interacting with the Heap is simple:
\begin{itemize}
  \item \texttt{malloc} - Allocates a chunk of memory that can store the desired size
  \item \texttt{free} - Mark an allocation as free for reuse
\end{itemize}

Although there are more functions that interact with the system Heap this paper mainly focuses on these main 2.

\subsubsection{malloc}
The C standard~\cite[p.~154-157]{ANSI_C} strictly defines the behavior of malloc and free. Upon successful execution, malloc provides a block of memory suitably aligned for any variable. The allocated memory is uninitialized, meaning it contains indeterminate values. If the function fails to allocate the requested memory, perhaps due to insufficient memory space, it returns a NULL pointer. Notably, the standard guarantees that malloc will not modify the allocated memory's content, differentiating it from similar functions like calloc, which sets the allocated memory to zero. The pointer returned by malloc is always a unique address not being used elsewhere in the program until it is freed by the \emph{free} function.

\subsubsection{free}
The \emph{free} function pretty much does the inverse of what malloc does. It marks the memory allocation starting at the address given as a parameter as free for use by the allocator. The algorithm behind the dynamic memory allocation API is left to be implementation-defined. We will dig deeper into the allocator used by LibC on the GNU/Linux operating system.

\subsection{Allocating memory in Linux}
The allocator used by the C standard library (GLibC) on Linux is a variation of \emph{ptmalloc}. There are two main ways to allocate memory using this allocator. The allocator automatically chooses which technique to use based on the size of the requested allocation\footnote{We will be referring to allocations with a size $<$ {M\_MMAP\_THRESHOLD} as \emph{small} and to allocations larger than that as \emph{big}}\footnote{{M\_MMAP\_THRESHOLD} is set to 127k bytes by default}:
\begin{itemize}
  \item \emph{small}~-~served by the allocation algorithm
  \item \emph{large}~-~served by a call to \emph{mmap}
\end{itemize}

Large allocations bypass the allocation algorithm entirely (and will not be the focus of our exploitation scenarios), unlike small allocations, which pass through a pretty complex set of steps before the memory is given to us. We will discuss this algorithm in detail, highlighting areas of interest that will help us develop exploits and mitigations later.

The allocation algorithm is described in depth in~\cite{MallocInternals}. We will mainly be exploring the \emph{tcache} part of the algorithm.

Definitions:
\begin{itemize}
  \item \emph{Arena}~-~An arena is a structure accessible by one or multiple threads, holding references to one or several heaps. It also includes linked lists of unallocated `chunks' within these heaps. Threads associated with a specific arena draw their memory from the free lists of that arena.
  \item \emph{Heap}~-~A heap is a sequential memory region divided into smaller units, known as chunks, for allocation purposes. Each heap is uniquely associated with a single arena.
  \item \emph{Chunk}~-~A chunk represents a small, allocatable memory section. This memory can be allocated (controlled by the application), freed (controlled by glibc), or merged with nearby chunks to create larger sections. Importantly, a chunk essentially serves as a shell encompassing the block of memory delivered to the application. Every chunk is part of one heap and linked to a single arena.
  \item \emph{Memory}~-~Memory refers to a segment of the application's address space, generally supported by RAM or swap
\end{itemize}

@@

  \item Part of the "top" chunk is split off, with a possibility of enlarging the "top" chunk prior to this.
\end{enumerate}

@@

\subsubsection{tcache}
Here is a more in-depth analysis of the \emph{tcache} functionality. This feature was introduced in GLibC in order to speed up the allocation of small, short-lived objects from multiple threads simultaneously. This part of the algorithm is the only one that is thread-local, the others requiring a lock on the process' Arena or atomic operations.

\texttt{tcache} intercepts the calls made to the \emph{free} function and stores these allocations in singly linked lists named tcache buckets. Each bucket can store a range of allocation sizes. When calling \emph{malloc} with tcache enabled, the first step is to check the appropriately sized tcache bucket for an allocation that might satisfy our requirements. If an allocation is found, it is removed from the singly linked list and returned.

\begin{figure}[ht]%
  \centering
  \begin{tikzpicture}[node distance=2cm]
    \tikzstyle{block} = [rectangle, draw, fill=blue!20, text width=5em, text
    centered, minimum height=2em, text width=2cm,
               text depth=0.25cm,
               text height=1em,
               align=center]
    \tikzstyle{line} = [draw, -latex']

    % Define matrix
    \matrix [matrix of nodes,
    nodes=block,
    column sep=0cm,
    row sep=0cm
    ] (m1) {
      $alloc_{0}$ & |[fill=white, draw=white]| \phantom{} & |[fill=white, draw=white]| \phantom{} \\
    };

    \matrix [matrix of nodes,
    below = 1cm of m1,
    nodes=block,
    column sep=0cm,
    row sep=0cm
    ] (m2) {
      $alloc_{1}$ & |[fill=white, draw=white]| \phantom{} & $alloc_{2}$ \\
    };

    \matrix [matrix of nodes,
    below = 1cm of m2,
    nodes=block,
    column sep=0cm,
    row sep=0cm
    ] (m3) {
      $bucket_{0}(2)$ & $bucket_{1}(0)$ & $bucket_{2}(1)$ \\
    };

    \draw[->] (m3-1-1) -- (m2-1-1);
    \draw[->] (m2-1-1) -- (m1-1-1);
    \draw[->] (m3-1-3) -- (m2-1-3);
  \end{tikzpicture}
  \caption{\label{fig:tcache} Sample tcache state}%
\end{figure}

The example in Figure~\ref{fig:tcache} is a sample tcache state after executing the following instruction sequqence:
\begin{enumerate}
  \item $free(alloc_{2})$
  \item $free(alloc_{0})$
  \item $free(alloc_{1})$
\end{enumerate}
Let's say we want to make an allocation that would fit inside $bucket_{0}$. The tcache algorithm would pull out that allocation from the list and return it to us, while updating the head of the list with the address of $alloc_{0}$.

\begin{figure}[ht]%
  \centering
  \begin{tikzpicture}[node distance=2cm]
    \tikzstyle{block} = [rectangle, draw, fill=blue!20, text width=5em, text
    centered, minimum height=2em, text width=2cm,
               text depth=0.25cm,
               text height=1em,
               align=center]
    \tikzstyle{line} = [draw, -latex']

    \matrix [matrix of nodes,
    below = 1cm of m1,
    nodes=block,
    column sep=0cm,
    row sep=0cm
    ] (m2) {
      $alloc_{0}$ & |[fill=white, draw=white]| \phantom{} & $alloc_{2}$ \\
    };

    \matrix [matrix of nodes,
    below = 1cm of m2,
    nodes=block,
    column sep=0cm,
    row sep=0cm
    ] (m3) {
      $bucket_{0}(1)$ & $bucket_{1}(0)$ & $bucket_{2}(1)$ \\
    };

    \draw[->] (m3-1-1) -- (m2-1-1);
    \draw[->] (m3-1-3) -- (m2-1-3);
  \end{tikzpicture}
  \caption{\label{fig:tcache} Sample tcache state after malloc}%
\end{figure}

How does \emph{tcache} store the links between the allocations?

\begin{figure}[ht]%
  \centering
  \begin{tikzpicture}[node distance=2cm]
    \tikzstyle{block} = [rectangle, draw, fill=blue!20, text width=5em, text
    centered, minimum height=2em, text width=2cm,
               text depth=0.25cm,
               text height=1em,
               align=center]
    \tikzstyle{line} = [draw, -latex']

    \matrix [matrix of nodes,
    below = 1cm of m1,
    nodes=block,
    column sep=0cm,
    row sep=0cm
    ] (m2) {
      fd & key & ... \\
    };

    \draw[->]  ([yshift=1cm]m2-1-1.north west)  -- node [midway, above, sloped, rotate=90, left=0.3cm] {start of alloc} (m2-1-1.north west);
  \end{tikzpicture}
  \caption{\label{fig:tcachechunk} Sample tcache entry}%
\end{figure}

The example in Figure~\ref{fig:tcachechunk} showcases how \emph{tcache} uses the allocated memory to store the required metadata to function. The value of the \emph{fd} contains the address of the next available allocation or \emph{NULL} if no other allocations are available. By default, each bucket contains a singly linked list of at most seven elements.

The \emph{key} value prevents freeing the same allocation twice. When the allocation is free, this value contains the address of the \emph{tcache\_perthread\_struct}. When the allocation is freed, this value is checked against the address of the per-thread struct, and if these values match, it means that the allocation has already been freed once, and the program aborts.

\subsubsection{Detecting heap corruption}
The \texttt{malloc} subsystem implements a series of measures to detect heap corruption throughout its codebase. Certain checks can consistently identify errors (for instance, passing an insufficiently aligned pointer to \texttt{free()}). However, most of these checks are heuristic and may not identify contrived, false chunks that mimic legitimate ones (for instance, checks for forward/backward linking of chunks). Consequently, heap corruption may persist undetected for an extended duration or not be reported.

Typical forms of corruption are managed via calls to \texttt{malloc\_printerr}; these checks are invariably included in the code. Additional checks utilize \texttt{assert} and can thus be disabled by building \texttt{glibc} with the \texttt{-DNDEBUG} flag. In the current \texttt{glibc} version, both types of checks terminate the process via a call to \texttt{\_\_libc\_message}, which ultimately invokes \texttt{abort}. Note that although old versions of \texttt{glibc} offered the ability to continue execution in the presence of heap corruption, this feature has since been discontinued.

@@

\section{Software Exploitation Techniques}%
Software exploitation has been a prevalent issue almost as early as the inception of software itself, underpinning the need for constant vigilance and innovation in software security. Early instances of software exploitation can be traced back to the late 1980s with the advent of the Morris Worm. Considered one of the first computer worms distributed via the Internet, the Morris Worm infected approximately 6,000 computers, causing significant slowdowns or rendering them unusable \cite{spafford1989internet}. Since then, various software vulnerabilities have been exploited, including buffer overflows, injection attacks, and privilege escalation, leading to massive data breaches, service disruptions, and significant financial losses \cite{owasp2017}. The history of software exploitation elucidates the perpetual arms race between attackers seeking to exploit vulnerabilities and defenders striving to secure systems. This dynamic continues to define the software security landscape today.

\subsection{Early stage}
\subsubsection{Shellcoding}
\begin{comment}
  shellcoding
  reference smashing the stack for fun and profit
\end{comment}
One of the earliest vulnerabilities that were exploited was the buffer overflow. A Buffer overflow represents a crucial class of software vulnerabilities that can serve as a launching pad for a myriad of complex exploits, one of which includes shellcode injection. This vulnerability arises when more data is written to a buffer than it can hold, allowing an attacker to overwrite adjacent memory locations \cite{seacord2013}. If exploited judiciously, this can alter the program's control flow, enabling an attacker to execute arbitrary code. One manifestation of this is shellcoding, where an attacker injects a sequence of bytes, often encoding command-line shell instructions, into the memory and manipulates the program's execution flow to execute this code \cite{one1996smashing}. This capability to execute arbitrary code gives the attacker complete control over the compromised system, illustrating the severity and implications of buffer overflow vulnerabilities.

@@

@@

But running the program with the following malicious crafted input we get a vastly different result\footnote{This would not happen in modern compilers because of mitigations enabled by default. The exact compilation flags required for this attack were: \texttt{-fno-stack-protector -z execstack -fno-pic -no-pie}}:
\begin{lstlisting}[caption={Shellcode execution}
      ,label={lst:bof_malicious}
      ,language=bash]
# The `cat` command is used just to keep
# the stdin of the `say_hi` program open
$ cat shellcode_malicious_input.bin - | ./say_hi
aaaa
Hi !
ls
Makefile  say_hi  shellcode_malicious_input.bin
\end{lstlisting}

\begin{lstlisting}[caption={Hexdump of malicious input}
      ,label={lst:bof_malicous_hex}
      ,language=bash]
0000000 0000 0000 0000 0000 0000 0000 0000 0000
*
0000100 0000 0000 0000 0000 115a 0040 0000 0000
0000110 9090 9090 9090 9090 9090 9090 9090 9090
*
0000310 8148 80ec 0000 6a00 4868 2fb8 6962 2f6e
0000320 2f2f 5073 8948 68e7 6972 0101 3481 0124
0000330 0101 3101 56f6 086a 485e e601 4856 e689
0000340 d231 3b6a 0f58 0005
0000347
\end{lstlisting}
Our simple program behaves like a system shell. We will now analyze what happened during the execution of our program. Here,~\ref {lst:bof_malicous_hex} is a hex dump of the input. Let us break it down:
\begin{enumerate}
  \item The first 0x100 bytes of the payload are just null bytes\footnote{We are allowed to do this because \emph{gets} does not stop reading the string when it encounters a \emph{NULL} (0) byte, even though strings in C are almost always NULL terminated.~\emph{scanf} will only return when it encounters an error or reads a line ending}. We do this in order to fill up the buffer that the application uses. These values could be anything (besides line endings).
  \item As shown in\ref{veziacistackuala} we now need to overflow the \emph{base pointer}. This exploitation scenario allows us to place any value instead of the saved rbp (because it is not used), but it can allow us to pivot the stack into some memory address we know in some cases.
  \item Now we overwrite the \emph{return address} to point to the address of the \emph{jump rsp} instruction.
  \item After the return address, we place a \emph{nop sled}. The purpose of the nop sled is to allow us some leeway\footnote{When jumping into shellcode written to the stack, the addresses the stack register and instruction pointer point to are close together. Stack manipulations (such as \emph{push}es and \emph{pop}s) can overwrite the shellcode, causing it to misbehave or making the processor execute illegal instructions.} when crafting the actual malicious payload.
  \item Finally, we place the instructions we want to execute. The payload exemplified\ref{vezipayload} here uses the \emph{execve} syscall to replace the current running process with the \emph{sh} shell process.
\end{enumerate}

This traditional exploitation technique of buffer overflow into shellcode has been effectively mitigated by a plethora of defensive mechanisms that we will discuss in subsequent chapters, such as Address Space Layout Randomization (ASLR), Data Execution Prevention (DEP), Stack Canaries, Control-Flow Integrity (CFI). Despite these advancements, the described exploitation scenarios retain relevance in contexts where these protection measures are absent or cannot be implemented. For example, systems lacking a Memory Management Unit (MMU), many embedded systems, and specific router configurations may not have these defenses, leaving them susceptible to such attacks. Thus, understanding these exploits remains critical in a comprehensive approach to system security.

@@

\subsubsection{Command injection}
Command injection vulnerabilities represent a pervasive class of security issues impacting web and system programming environments. These vulnerabilities occur when an application passes unsafe user-supplied data (forms, cookies, HTTP headers) to a system shell, allowing the attacker to execute arbitrary commands on the host operating system. In web applications, these flaws are primarily found in scripts that fail to properly sanitize user input, making them susceptible to cross-site scripting (XSS) and SQL injection attacks. However, command injection vulnerabilities are not confined to web programming; they are equally prevalent in systems programming. Poorly designed system applications and scripts that accept untrusted input for command execution can expose the system to similar attacks. This vulnerability, thus, underlines the importance of careful data handling and proper input validation, irrespective of the programming environment.
